Equalizer\\Programming and User Guide

### 2013-07-25 Developer/ProgrammingGuide/paper.tex
Programming Guide for Equalizer 1.6

\date{
  \vspace{14cm}
  \begin{textblock}{}(0,4.5)
    \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{textblock}
      {\Large The official reference for developing and
        deploying parallel, scalable OpenGL\texttrademark\ applications using
        the Equalizer parallel rendering framework}\\\vspace{1cm}
      Version 1.12 for Equalizer 1.4\\[\medskipamount]
  \today
}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\sref}[1]{Section~\ref{#1}}
\newcommand{\aref}[1]{Appendix~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}
\indentcaption{2em}
\setcounter{tocdepth}{2}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
    1.10    & Feb 17, 2012 & Revision for Equalizer 1.2\\
    1.12    & Jul 20, 2012 & Revision for Equalizer 1.4\\
    1.11.1  & Mar 20, 2013 & Adapt to new head tracking event\\
    \hline
  \end{tabularx}}
\clearpage

@@

Equalizer can be installed by downloading a binary installer or source
code\footnote{\link{http://www.equalizergraphics.com/downloads.html}}. After
installing Equalizer, please take a look at the Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
to get familiar with the capabilities of Equalizer and the \textsf{eqPly}
example.

Equalizer uses CMake to generate platform-specific build files. Compiling
Equalizer from source is as simple as running \textsf{make} on Linux or Mac OS
X, running \textsf{make xcode} and building the generated XCode project on Mac
OS X, or creating a Equalizer Visual Studio solution using \textsf{CMake} on
Windows.

@@

\subsubsection{Server}
The Equalizer server is responsible for managing one visualization session on a
shared memory system or graphics cluster. Based on its configuration and
controlling input from the application, it computes the active resources,
updates the configuration and generates tasks for all processes. Furthermore it
controls and launches the application's rendering client processes. The
Equalizer server is the entity in charge of the configuration, and all other
processes receive their configuration from the server.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/2D.pdf}
  {\caption{2D Compound}}
\end{wrapfigure}
2D decomposes the rendering in screen-space, that is, each contributing
rendering unit processes a tile of the final view. The recomposition
simply assembles the tiles side-by-side on the destination view. This
mode is also known as sort-first or SFR.

The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the amount of pixel data for the destination view.

@@

DB, as shown in \fig{fDB}\footnote{3D model courtesy of AVS, USA.}, decomposes
the rendered database so that all rendering units process a part of the scene in
parallel. This mode is also known as sort-last, and is very similar to the data
decomposition approach used by HPC applications.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DB.pdf}
  {\caption{\label{fDB}Database Compound}}
\end{wrapfigure}
Volume rendering applications use an ordered alpha-based blending to
composite the result image. The depth buffer information is used to
composite the individual images correctly for polygonal data.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}
  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
  {\caption{A DPlex Compound}}
\end{wrapfigure}
DPlex compounds assign full, alternating frames to individual rendering
units. The resulting images are copied to the destination channel, and
Equalizer load-balancing is used to ensure a steady framerate on the
destination window. This mode is also known as time-multiplex or AFR.

Due to the frame consistency between consecutive frames, this mode scales
very well. The IO requirements for pixel transfer are small and
constant.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}Pixel Compound}}\vspace{-8ex}
\end{wrapfigure}
Pixel compounds are similar to 2D compounds. While 2D compounds decompose the
screen space into contiguous regions, pixel compounds assign one pixel of a
regular kernel to each resource. The frusta of the source rendering units are
modified so that each unit renders an evenly distributed subset of pixels, as
shown in \fig{fPixel}\footnote{3D model courtesy of AVS, USA.}

As 2D compounds, pixel compounds have low, constant IO requirements for the
pixel transfers during recomposition. They are naturally load-balanced for
fill-limited operations, but do not scale geometry processing at all.

@@

\begin{wrapfigure}{l}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
  {\caption{\label{fPixelKernel}\small Pixel Compound Kernel}}
\end{wrapfigure}
Pixel compounds work well for purely fill-limited applications. Techniques like
view frustum culling do not reduce the rendered data, since each resource has
approximately the same frustum. Pixel compounds are ideal for ray-tracing, which
is highly fill-limited and needs the full database for rendering anyway. Volume
rendering applications are also well suited for this mode, and should choose it
over 2D compounds.

Pixel compounds in Equalizer are configured using the \textsf{pixel}
parameter, using the values \textsf{[ x y width height]}  to configure
the size and offset of the sampling kernel. The width and height of the
sampling kernel define how many pixels are skipped in the x and y
direction, respectively. The x and y offset define the index of the
source channel within the kernel. They have to be smaller than the size
of the kernel. \fig{fPixelKernel} illustrates these parameters, and
\fig{fPixelKernels} shows some example kernels for a four-to-one pixel decomposition.

@@

Applications can use subpixel compounds to accelerate depth-of-field effects and
software anti-aliasing, with potentially a different number of idle or non-idle
samples per pixel.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Subpixel.pdf}
  {\caption{\label{fSubpixel}Subpixel Compound}}
\end{wrapfigure}
Subpixel compounds are configured using the \textsf{subpixel} parameter, using
the values \textsf{[~index size ]}. The \textsf{index} parameter corresponds to
the current resource index and the \textsf{size} is the total number of
resources used.

@@

Multiple channels in application windows may be used to view the model
from different viewports.

Sometimes, a single window is split across multiple projectors, e.g., by
using an external splitter such as the Matrox TripleHead2Go. In this
case configure one channel for each segment, using the channel's
\textsf{viewport} to configure its position relative to the window.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and Projection Parameters}}
\end{wrapfigure}
If you are writing a configuration for workstation usage you can skip
the following sections and restart with \sref{sCompounds}.

@@

The frustum can be specified as a wall or projection description. Take
care to choose your reference system for describing the frustum to be
the same system as used by the head-tracking matrix calculated by the
application.

A wall is completely defined by the bottom-left, bottom-right and
top-left coordinates relative to the origin.

A projection is defined by the position and head-pitch-roll orientation
of the projector, as well as the horizontal and vertical field-of-view
and distance of the projection wall.

\fig{fFrusta} illustrates the wall and projection frustum parameters.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
  {\caption{\label{fCanvas}A Canvas using four Segments}}
\end{wrapfigure}
To synchronize the video output, configure either a canvas swap barrier or a
swap barrier on each segment to be synchronized.

@@

A segment represents one output channel of the canvas, e.g., a projector
or an LCD. A segment has an output channel, which references the channel to
which the display device is connected.

@@

Configure one \textsf{view} for each logical view in each layout. Set the
\textsf{viewport} to position the view. Set the mode to stereo if appropriate.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}Layout with four Views}}
\end{wrapfigure}
A view has a fractional viewport relative to its layout.  A layout
is often fully covered by its views, but this is not a requirement.

@@

A view might have an observer, in which case its frustum is tracked by
this observer.

\fig{fLayout} shows an example layout using four views on a single
segment. \fig{fDisplay} shows a real-world setup of a single canvas with six
segments using underlap, with a two-view layout activated. This configuration
generates eight destination channels.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small 2D Load-Balancing}}
\end{wrapfigure}
Equalizers are used to update compound parameters based on runtime
data. They are attached to a compound (sub-)tree, on which they
operate. The Equalizer distribution contains numerous example
configuration files using equalizers.

@@

Using a load equalizer is transparent to the application, and can be used with
any application for 2D, and with most applications for DB load-balancing. Some
applications do not support dynamic updates of the database range, and therefore
cannot be used with DB load-balancing.

@@

Depending on the model position and data structure, each segment of a
multi-display system has a different rendering load. The segment with
the biggest load determines the overall performance when using a static
assignment of resources to segments. The view equalizer analyzes the
load of all segments, and adjusts the resource usage each frame. It
equalizes the load on all segments of a view.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
      Load-Balancing for two Segments using eight GPUs}}
\end{wrapfigure}
\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
  AG, 2008} illustrates this process. On the left side, a static
assignment of resources to display segments is used. The right-hand
segment has a higher load than the left-hand segment, causing
sub-optimal performance. The configuration on the right uses a view
equalizer, which assigns two GPUs to the left segment and four GPUs to
the right segment, which leads to optimal performance for this model and
camera position.

The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particularly useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

\begin{wrapfigure}{l}{.382\textwidth}
    \includegraphics[width=.382\textwidth]{images/viewLB.png}
    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
        Load-Balancing for a CAVE}}
\end{wrapfigure}
\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
by five GPUs. The front and left segments show the model and have a
significant rendering load. The view equalizer assigns the GPUs from
the top, bottom and right wall for rendering the left and front wall in
this configuration.

@@

Dynamic Frame Resolution (DFR) trades rendering performance for visual
quality. The rendering for a channel is done at a different resolution
than the native channel resolution to keep the framerate constant. The
DFR equalizer adjusts the zoom of a channel, based on the target and
current framerate. It is typically used for fill-rate bound
applications, such as volume rendering and ray-tracing.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}Dynamic Frame Resolution}}
\end{wrapfigure}
\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

@@

@@

\section{\label{sHello}Hello, World!}

The \textsf{eqHello} example is a minimal application to illustrate the basic
principle of any Equalizer application: The application developer has to
implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
applications.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/eqHello.png}
  {\caption{\label{fHello}Hello, World!}}
\end{wrapfigure}
It can be run as a stand-alone application from the command line or using an
explicit configuration file for a visualization cluster. In the stand-alone
case, any Equalizer application, including eqHello, will automatically configure
itself to use all graphics cards found on the local system for scalability.

@@

\section{Programming Interface}

The API documentation is available on the website or in the header files, and
provides a comprehensive documentation on individual methods, types and other
elements of the API. Methods marked with a specific version are part of the
official, public API and have been introduced by this Equalizer
version. Reasonable care is taken to not break API compatibility or change the
semantics of these methods within future Equalizer versions of the same major
revision. Any eventual changes to the public API are documented in the release
notes and the file \textsf{CHANGES.txt}.

@@

The Equalizer software stack is modularized, layering gradually more powerful,
but less generic APIs on to of each other. It furthermore relies on a number of
required and optional libraries. Application developers are exposed to the
following name\-spa\-ces:

@@

\section{\label{sSeqPly}Sequel: The seqPly Polygonal Renderer}

@@

\section{\label{sEqPly}Equalizer: The eqPly Polygonal Renderer}

In this section the source code of \textsf{eqPly} is used to introduce the
Equalizer API in detail, and relevant design decisions, caveats and other
remarks are discussed.

\textsf{eqPly} is a parallel renderer for polygonal data in the
\textsf{ply} file format. It supports all relevant Equalizer features,
and can be used to render on large-scale displays, immersive
environments with head tracking and to render massive data sets using
all scalable rendering features of Equalizer. It is a superset of
\textsf{seqPly}, introduced in \sref{sSeqPly}.

@@

The distributed data classes are helper classes based on
\textsf{co::Serializable} or its base class \textsf{co::Object}. They illustrate
the typical usage of distributed objects for static as well as dynamic,
frame-specific data. Furthermore they are used for a basic scene graph
distribution of the model data to the render client processes.

@@

\subsubsection{Main Loop}

@@

@@

\subsection{\label{sNetObject}Distributed Objects}

@@

\subsubsection{\label{sInitData}InitData - a Static Distributed Object}

@@

{\footnotesize\lstinputlisting[firstline=66, lastline=77]
  {../../../Equalizer/examples/eqPly/initData.cpp}}

\subsubsection{FrameData - a Versioned Distributed Object}

@@

{\footnotesize\lstinputlisting[firstline=123, lastline=130]
  {../../../Equalizer/examples/eqPly/frameData.h}}
{\footnotesize\lstinputlisting[firstline=54, lastline=82]
  {../../../Equalizer/examples/eqPly/frameData.cpp}}

\subsubsection{\label{sSceneData}Scene Data}

@@

\subsection{Config}

@@

\subsubsection{Initialization and Exit}

@@

The exit function of the configuration stops the render clients by calling
\textsf{eq::Con\-fig::exit}, and then de-registers the initialization and
frame data objects with the session:

@@

\subsubsection{Frame Control}

The rendering frames are issued by the application. The \textsf{eqPly::Config}
overrides \textsf{startFrame} to update its data, commit a new version of the
frame data object, and then requests the rendering of a new frame using the
current frame data version. This version is passed to the rendering callbacks
and will be used by the rendering threads to synchronize the frame data to the
state belonging to the current frame:

{\footnotesize\lstinputlisting[firstline=294, lastline=301]
  {../../../Equalizer/examples/eqPly/config.cpp}}

Updating the shared data is decomposed in different steps. If a tracker is used,
the current head position and orientation is retrieved and passed to Equalizer,
which uses the head matrix together with the wall or projection description to
compute the view frusta, as explained in \sref{sTracking}:

{\footnotesize\lstinputlisting[firstline=303, lastline=311]
  {../../../Equalizer/examples/eqPly/config.cpp}}

The camera position is recomputed based on the current navigation mode, and the
idle state for rendering is set. When idle, \textsf{eqPly} performs
anti-aliasing to gradually reduce aliasing effects in the rendering. The idle
state is tracked by the application and used by the rendering callbacks to
jitter the frusta, accumulate and display the results:

{\footnotesize\lstinputlisting[firstline=313, lastline=343]
  {../../../Equalizer/examples/eqPly/config.cpp}}

\subsubsection{Event Handling}

Events are sent by the render clients to the application using
\textsf{eq::Config::sendEvent}. At the end of the frame,
\textsf{Config::finishFrame} calls \textsf{Config::handleEvents} to do
the event handling. The default implementation processes all pending
events by calling \textsf{Config::handleEvent} for each of them.

Since \textsf{eqPly} uses event-driven execution, the config maintains a
dirty state to allow the application to know when a redraw is needed.

@@

\subsubsection{Initialization and Exit}

>>>>

\date{
  \vspace{14cm}
  \begin{textblock}{}(0,4.5)
    \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{textblock}
      {\Large The official reference for developing and
        deploying parallel, scalable OpenGL\texttrademark\ applications using
        the Equalizer parallel rendering framework}\\\vspace{1cm}
      Version 1.14 for Equalizer 1.6\\[\medskipamount]
  \today
}

\newcommand{\tm}{\texttrademark~}
\newcommand{\rc}{\raise 1ex\hbox{{\tiny\textregistered}}~}
\newcommand{\fig}[1]{Figure~\ref{#1}}
\newcommand{\sref}[1]{Section~\ref{#1}}
\newcommand{\aref}[1]{Appendix~\ref{#1}}
\newcommand{\link}[1]{\htmladdnormallink{#1}{#1}}
\indentcaption{2em}
\setcounter{tocdepth}{3}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
    1.10    & Feb 17, 2012 & Revision for Equalizer 1.2\\
    1.12    & Jul 20, 2012 & Revision for Equalizer 1.4\\
    1.14    & Jul 25, 2013 & Revision for Equalizer 1.6\\
    \hline
  \end{tabularx}}
\clearpage

@@

Equalizer can be installed by using a package or building the source
code\footnote{\link{http://www.equalizergraphics.com/downloads.html}}. After
installing Equalizer, please take a look at the Quickstart
Guide\footnote{\link{http://www.equalizergraphics.com/documents/EqualizerGuide.html}}
to get familiar with the capabilities of Equalizer and the \textsf{eqPly}
example. Currently we provide Ubuntu packages on
launchpad\footnote{\link{https://launchpad.net/~eilemann/+archive/equalizer}}
and MacPorts portfiles\footnote{\link{https://github.com/Eyescale/portfiles}}.

Equalizer uses Buildyard and cmake to generate platform-specific build
files. Compiling Equalizer from source consist of cloning Buildyard, the
Eyescale configurations and then building Equalizer:

{\footnotesize\begin{lstlisting}
  git clone https://github.com/Eyescale/Buildyard.git
  cd Buildyard
  git clone https://github.com/Eyescale/config.git config.eyescale
  make Equalizer
\end{lstlisting}}

The Windows build is similar, except that CMake will generate a Visual Studio
solution which is used to build Equalizer.

@@

\subsubsection{Server}
The Equalizer server is responsible for managing one visualization session on a
shared memory system or graphics cluster. Based on its configuration and
controlling input from the application, it computes the active resources,
updates the configuration and generates tasks for all processes. Furthermore it
controls and launches the application's rendering client processes. The
Equalizer server is the entity in charge of the configuration, and all other
processes receive their configuration from the server. It typically runs as a
separate entity within separate threads in the application process.

@@

2D decomposes the rendering in screen-space, that is, each contributing
rendering unit processes a tile of the final view. The recomposition
simply assembles the tiles side-by-side on the destination view. This
mode is also known as sort-first or SFR.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/2D.pdf}
  {\caption{2D Compound}}
\end{wrapfigure}
The advantage of this mode is a low, constant IO overhead for the pixel
transfers, since only color information has to be transmitted. The upper
limit is the amount of pixel data for the destination view.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DB.pdf}
  {\caption{\label{fDB}Database Compound}}
\end{wrapfigure}
DB, as shown in \fig{fDB}\footnote{3D model courtesy of AVS, USA.}, decomposes
the rendered data\-base so that all rendering units process a part of the scene
in parallel. This mode is also known as sort-last, and is very similar to the
data decomposition approach used by HPC applications.

Volume rendering applications use an ordered alpha-based blending to
composite the result image. The depth buffer information is used to
composite the individual images correctly for polygonal data.

@@

DPlex compounds assign full, alternating frames to individual rendering
units. The resulting images are copied to the destination channel, and
Equalizer load-balancing is used to ensure a steady framerate on the
destination window. This mode is also known as time-multiplex or AFR.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/DPlex.pdf}
  {\caption{A DPlex Compound}}
\end{wrapfigure}
Due to the frame consistency between consecutive frames, this mode scales
very well. The IO requirements for pixel transfer are small and
constant.

@@

Pixel compounds are similar to 2D compounds. While 2D compounds decompose the
screen space into contiguous regions, pixel compounds assign one pixel of a
regular kernel to each resource. The frusta of the source rendering units are
modified so that each unit renders an evenly distributed subset of pixels, as
shown in \fig{fPixel}\footnote{3D model courtesy of AVS, USA.}

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}Pixel Compound}}
\end{wrapfigure}
As 2D compounds, pixel compounds have low, constant IO requirements for the
pixel transfers during recomposition. They are naturally load-balanced for
fill-limited operations, but do not scale geometry processing at all.

@@

Pixel compounds work well for purely fill-limited applications. Techniques like
view frustum culling do not reduce the rendered data, since each resource has
approximately the same frustum. Pixel compounds are ideal for ray-tracing, which
is highly fill-limited and needs the full database for rendering anyway. Volume
rendering applications are also well suited for this mode, and should choose it
over 2D compounds.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/pixelKernel.pdf}
  {\caption{\label{fPixelKernel}\small Pixel Compound Kernel}}
\end{wrapfigure}
Pixel compounds in Equalizer are configured using the \textsf{pixel}
parameter, using the values \textsf{[ x y width height]}  to configure
the size and offset of the sampling kernel. The width and height of the
sampling kernel define how many pixels are skipped in the x and y
direction, respectively. The x and y offset define the index of the
source channel within the kernel. They have to be smaller than the size
of the kernel. \fig{fPixelKernel} illustrates these parameters, and
\fig{fPixelKernels} shows some example kernels for a four-to-one pixel decomposition.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/Subpixel.pdf}
  {\caption{\label{fSubpixel}Subpixel Compound}}
\end{wrapfigure}
Applications can use subpixel compounds to accelerate depth-of-field effects and
software anti-aliasing, with potentially a different number of idle or non-idle
samples per pixel.

@@

Subpixel compounds are configured using the \textsf{subpixel} parameter, using
the values \textsf{[~index size ]}. The \textsf{index} parameter corresponds to
the current resource index and the \textsf{size} is the total number of
resources used.

@@

Multiple channels in application windows may be used to view the model from
different viewports. Sometimes, a single window is split across multiple
projectors, e.g., by using an external splitter such as the Matrox
TripleHead2Go. In this case configure one channel for each segment, using the
channel's \textsf{viewport} to configure its position relative to the window.

@@

If you are writing a configuration for workstation usage you can skip
the following sections and restart with \sref{sCompounds}.

@@

The frustum can be specified as a wall or projection description. Take care to
choose your reference system for describing the frustum to be the same system as
used by the head-tracking matrix calculated by the application.  A wall is
completely defined by the bottom-left, bottom-right and top-left coordinates
relative to the origin.  A projection is defined by the position and
head-pitch-roll orientation of the projector, as well as the horizontal and
vertical field-of-view and distance of the projection wall.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and Projection Parameters}}
\end{wrapfigure}
\fig{fFrusta} illustrates the wall and projection frustum parameters.

@@

To synchronize the video output, configure either a canvas swap barrier or a
swap barrier on each segment to be synchronized.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
  {\caption{\label{fCanvas}A Canvas using four Segments}}
\end{wrapfigure}
A segment represents one output channel of the canvas, e.g., a projector
or an LCD. A segment has an output channel, which references the channel to
which the display device is connected.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}Layout with four Views}}\vspace{-6ex}
\end{wrapfigure}
Configure one \textsf{view} for each logical view in each layout. Set the
\textsf{viewport} to position the view. Set the mode to stereo if appropriate.

@@

A view has a fractional viewport relative to its layout.  A layout
is often fully covered by its views, but this is not a requirement.

@@

A view might have an observer, in which case its frustum is tracked by this
observer. \fig{fLayout} shows an example layout using four views on a single
segment. \fig{fDisplay} shows a real-world setup of a single canvas with six
segments using underlap, with a two-view layout activated. This configuration
generates eight destination channels.

@@

Equalizers are used to update compound parameters based on runtime
data. They are attached to a compound (sub-)tree, on which they
operate. The Equalizer distribution contains numerous example
configuration files using equalizers.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/lb.pdf}
  {\caption{\label{fLoadBalancing}\small 2D Load-Balancing}}
\end{wrapfigure}
Using a load equalizer is transparent to the application, and can be used with
any application for 2D, and with most applications for DB load-balancing. Some
applications do not support dynamic updates of the database range, and therefore
cannot be used with DB load-balancing.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-Segment
      Load-Balancing for two Segments using eight GPUs}}
\end{wrapfigure}
Depending on the model position and data structure, each segment of a
multi-display system has a different rendering load. The segment with
the biggest load determines the overall performance when using a static
assignment of resources to segments. The view equalizer analyzes the
load of all segments, and adjusts the resource usage each frame. It
equalizes the load on all segments of a view.

\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
  AG, 2008} illustrates this process. On the left side, a static
assignment of resources to display segments is used. The right-hand
segment has a higher load than the left-hand segment, causing
sub-optimal performance. The configuration on the right uses a view
equalizer, which assigns two GPUs to the left segment and four GPUs to
the right segment, which leads to optimal performance for this model and
camera position.

\begin{wrapfigure}{r}{.382\textwidth}
    \includegraphics[width=.382\textwidth]{images/viewLB.png}
    {\caption{\label{fViewLoadBalancing2}\small Cross-Segment
        Load-Balancing for a CAVE}}
\end{wrapfigure}
The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particularly useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

\fig{fViewLoadBalancing2} shows cross-usage for a five-sided CAVE driven
by five GPUs. The front and left segments show the model and have a
significant rendering load. The view equalizer assigns the GPUs from
the top, bottom and right wall for rendering the left and front wall in
this configuration.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}Dynamic Frame Resolution}}
\end{wrapfigure}
Dynamic Frame Resolution (DFR) trades rendering performance for visual
quality. The rendering for a channel is done at a different resolution
than the native channel resolution to keep the framerate constant. The
DFR equalizer adjusts the zoom of a channel, based on the target and
current framerate. It is typically used for fill-rate bound
applications, such as volume rendering and ray-tracing.

\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

@@

This Programming Guide introduces Equalizer using a top-down approach, starting
with a general introduction of the API in \sref{sAPI}, followed by the
simplified Sequel API in \sref{sSeqPly} which implements common use cases to
deliver a large subset of the canonical Equalizer API introduced in
\sref{sEqualizer}. \sref{sNetwork} introduces the separate Collage network
library used as the foundation for the distributed execution and data
synchronizing throughout Sequel and Equalizer.

\section{\label{sAPI}Programming Interface}

@@

The API documentation is available on the website or in the header files, and
provides a comprehensive documentation on individual methods, types and other
elements of the API. Methods marked with a specific version are part of the
official, public API and have been introduced by this Equalizer
version. Reasonable care is taken to not break API compatibility or change the
semantics of these methods within future Equalizer versions of the same major
revision. Any changes to the public API are documented in the release notes and
the file \textsf{CHANGES.txt}.

\subsection{\label{sHello}Hello, World!}

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/eqHello.png}
  {\caption{\label{fHello}Hello, World!}}
\end{wrapfigure}
The \textsf{eqHello} example is a minimal application to illustrate the basic
principle of any Equalizer application: The application developer has to
implement a rendering method similar to the \textsf{glutDisplayFunc} in GLUT
applications.

It can be run as a stand-alone application from the command line or using an
explicit configuration file for a visualization cluster. In the stand-alone
case, any Equalizer application, including eqHello, will automatically configure
itself to use all graphics cards found on the local system for scalability.

@@

@@

The Equalizer software stack is modularized, layering gradually more powerful,
but less generic APIs on top of each other. It furthermore relies on a number of
required and optional libraries. Application developers are exposed to the
following name\-spa\-ces:

@@

\section{\label{sSeqPly}The Sequel Simple Equalizer API}

@@

\section{\label{sEqualizer}The Equalizer Parallel Rendering Framework}

The core Equalizer client library exposes the full feature set and flexibility
to the application developer. New application development is encouraged to use
the Sequel library, and inquire on the \textsf{eq-dev} mailing list when
specific functionality is not available in Sequel. Only advanced and complex
applications should be implemented using the Equalizer client library directly.

\subsection{\label{sEqPly}The eqPly Polygonal Renderer}
In this section the source code of \textsf{eqPly} is used to introduce the
Equalizer API in detail, and relevant design decisions, caveats and other
remarks are discussed.

\textsf{eqPly} is a parallel renderer for polygonal data in the \textsf{ply}
file format. It supports nearly all Equalizer features, and can be used to
render on large-scale displays, immersive environments with head tracking and to
render massive data sets using all scalable rendering features of Equalizer. It
is a superset of \textsf{seqPly}, introduced in \sref{sSeqPly}.

@@

The distributed data classes are helper classes based on
\textsf{co::Serializable} or its base class \textsf{co::Object}. They illustrate
the typical usage of distributed objects for static as well as dynamic,
frame-specific data. Furthermore they are used for a basic scene graph
distribution of the model data to the render client processes. \sref{sNetwork}
provides more background on the Collage network library.

\subsubsection{main Function}

@@

\paragraph{Main Loop}

@@

\paragraph{Render Clients}

@@

\subsubsection{\label{sNetObject}Distributed Objects}

@@

\paragraph{\label{sInitData}InitData - a Static Distributed Object}

@@

{\footnotesize\lstinputlisting[firstline=62, lastline=73]
  {../../../Equalizer/examples/eqPly/initData.cpp}}

\paragraph{FrameData - a Versioned Distributed Object}

@@

{\footnotesize\lstinputlisting[firstline=124, lastline=131]
  {../../../Equalizer/examples/eqPly/frameData.h}}
{\footnotesize\lstinputlisting[firstline=53, lastline=81]
  {../../../Equalizer/examples/eqPly/frameData.cpp}}

\paragraph{\label{sSceneData}Scene Data}

@@

\subsubsection{Config}

@@

\paragraph{Initialization and Exit}

@@

The exit function of the configuration stops the render clients by calling
\textsf{eq::Con\-fig::exit}, and then de-registers the initialization and
frame data objects:

@@

\paragraph{Frame Control}

The rendering frames are issued by the application main loop. The
\textsf{eqPly::Config} overrides \textsf{startFrame} to update its data, commit
a new version of the frame data object, and then requests the rendering of a new
frame using the current frame data version. This version is passed to the
rendering callbacks and will be used by the rendering threads to synchronize the
frame data to the state belonging to the current frame. This ensures that all
frame-specific data, e.g., the camera position, is used consistently to generate
the frame:

{\footnotesize\lstinputlisting[firstline=274, lastline=281]
  {../../../Equalizer/examples/eqPly/config.cpp}}

The update of the per-frame shared data consist of calculating the camera
position based on the current navigation mode, and determining the idle state
for rendering. When idle, \textsf{eqPly} performs anti-aliasing to gradually
reduce aliasing effects in the rendering. The idle state is tracked by the
application and used by the rendering callbacks to jitter the frusta, accumulate
and display the results, as described in \sref{sFSAA}:

{\footnotesize\lstinputlisting[firstline=283, lastline=315]
  {../../../Equalizer/examples/eqPly/config.cpp}}

\paragraph{Event Handling}

Events are sent by the render clients to the application using
\textsf{eq::Config::sendEvent}. At the end of the frame,
\textsf{Config::finishFrame} calls \textsf{Config::handle\-Events} to perform
the event handling. The default implementation processes all pending events by
calling \textsf{Config::handleEvent} for each of them.

Since \textsf{eqPly} uses event-driven execution, the config maintains a dirty
state to know when a redraw is needed.

@@

\paragraph{Initialization and Exit}
